---
title: "04_popgen"
format: html
editor: visual
---

## .vcf files from single nucleotide variant calling with ref_map.pl in STACKS

```{r}
library(SNPfiltR)
library(vcfR)
```

```{bash, eval = FALSE}
#transfer vcf file from hyak to raven
rsync --archive --progress --verbose lbreit@klone.hyak.uw.edu:/gscratch/scrubbed/lbreit/fish546/REFMAP/test2/populations.snps.vcf ../data

#transfer pop map from hyak to raven
rsync --archive --progress --verbose lbreit@klone.hyak.uw.edu:/mmfs1/home/lbreit/fish546/popmap_test2.txt ../data
```

```{r}
#read vcf into this R session as vcfR 
vcfR <- read.vcfR("~/fish546/lara-zostera/project/data/populations.snps.vcf")

# check metadata present in vcf
vcfR

head(vcfR, 3)
```

Understanding a VCF file:

**Fixed section** corresponds to per-variant genotype information

-   CHROM: the chromosome where the variant is located (zm has 6 pairs of chromosomes)
-   POS: the 1-based position of the variant on the chromosome
-   ID: a unique identifier for the variant. *unsure how to read format*
-   REF: reference allele
-   ALT: alternate allele, the one that is observed as a variant in this position
-   QUAL: quality score. "NA" indicates either missing or unreported quality scores *ask Bryan about this*
-   FILTER: filter status for this variant. "PASS" indicates that the variant passed filtering criteria during variant calling.

**Genotype section** corresponds to sample-specific data columns of a VCF file; i.e., each sample's genotype at each variant site

Sample size: `r length(colnames(vcfR@gt)[-1])` individuals

Variants: `r nrow(vcfR@fix)` biallelic sites

AHHH!!!!! IT WORKED!!!!

```{r}
#pop map
#read all lines from the file
lines <- readLines("../data/popmap_test2.txt")

# Filter out lines that start with '#'
lines <- lines[!grepl("^#", lines)]

# Convert the remaining lines into a textConnection for reading as a table
popmap <- read.table(text = lines, sep = "\t", header = FALSE)

# Assign column names
colnames(popmap) <- c("id", "pop", "region")
head(popmap)
```

## Why filter?

We want to detect and and remove markers that do not actually represent loci, identify and correct erroneous SNP calls, and assess genotyping error. Error can be introduced during library prep and bioinformatic processing! See O'Leary et al. (2018) **Table 1** for a summary of RADseq issues, technical and bioinformatic mitigation (filtering) of those issues.

## hard filtering

```{r}
#start by visualizing the distributions of depth of sequencing and genotype quality among called genotypes

#these histograms represent the frequency of a particular genotype depth (DP) or genotype quality (GQ) value summed over every sample at every SNP position in the dataset

hard_filter(vcfR=vcfR)
```

The histogram of dp.matrix shows the full genoptype depth across all genotypes. A very large spike at low depth, and a long tail extending to very high depths (up to 20,000+ reads). Dashed red line is mean depth. Extremely high-depth genotypes may suggest problematic loci. Set both a min and a max for filtering.

The histogram of dp.matrix\[dp.matrix \<25\] shows the distribution of genotype depth for depths less than 25 (close up of the left side of the first histogram). Most genotypes have low sequencing depth. We will proceed with a depth cut-off of 10 (standard), under which genotype calls are unreliable.

The histogram of gq.matrix shows the distribution of genotype quality (GQ) scores across all genotypes in the dataset. The plot is heavily skewed towards high GQ values. Red dashed line represents mean. Many workflows will remove genotypes with GQ below 30, as lower values indicate less confidence in genotype call.

### genotype depth ≥ 10 and gq \> 30

```{r}
vcfR<-hard_filter(vcfR=vcfR, depth = 10, gq = 30)
```

### genotype allelic balance \< 0.2 and \> 0.8

If a SNP is observed in only one or two readouts of many (e.g., \>10), it is likely a sequencing error.

```{r}
# Allelic balance can be measured by comparing the number of reads supporting each allele at a specific locus. 
vcfR.AB <- filter_allele_balance(vcfR, min.ratio = 0.2, max.ratio = 0.8)
```

Still don't quite understand why we filter for allelic imbalance. I know that ...

### remove monomorphic SNPs

```{r}
# filters out non-informative SNPs. A minor allele count (MAC) of 0 provides no genetic variation for analysis

vcfR.depth.AB.mac <- min_mac(vcfR.AB, min.mac = 1) #would we want the MAC threshold to be higher to reduce false positives?
```

**after hard filtering**

Sample size: `r length(colnames(vcfR.depth.AB.mac@gt)[-1])` individuals

Variants: `r nrow(vcfR.depth.AB.mac@fix)` SNPs

## soft filtering

```{r soft_filtering_1, echo=TRUE, message=FALSE, warning=FALSE}

# Remove LOCI with > 20% missing data
vcfR.depth.AB.mac.snp <- missing_by_snp(vcfR.depth.AB.mac, cutoff = 0.5)

# Keep one SNP per RAD
vcfR_maf <- vcfR::maf(vcfR.depth.AB.mac.snp) # calculate MAF per locus
vcfR_maf_df <- as.data.frame(vcfR_maf) # transform to data frame for manipulation
vcfR_maf_df$Tag <- sub("\\:.*", "", rownames(vcfR_maf)) # separate tag IDs
vcfR_maf_df$Number <- 1:length(vcfR_maf_df$Tag) # just number all the rows

vcfR_maf_df_sum <- vcfR_maf_df %>%
  tibble::rownames_to_column('Site') %>% # turn row names into a variable
  dplyr::group_by(Tag) %>% # group observations by Tag ID
  dplyr::slice(which.max(Frequency)) %>% # keep observations with highest maf
  dplyr::arrange(Number) # sort by increasing

vcfR.depth.AB.mac.snp.thin <- vcfR.depth.AB.mac.snp[c(vcfR_maf_df_sum$Number),] # subset those observations from the previous vcf

head(vcfR.depth.AB.mac.snp.thin)
```
*Why are we thinning to one SNP per RAD tag?*

Multiple SNPs within RAD tags can complicate linkage disequilibrium (LD) analyses, as SNPs from the same tag are physically linked. Pipelines often thin SNPs to one SNP per RAD tag (as we've done here). Retaining the SNP with the highest minor allele frequency (MAF) ensures genetic informativeness while minimizing LD bias.

Pre-thinning: `r nrow(vcfR.depth.AB.mac.snp@fix)` SNPs

Post-thinning: `r nrow(vcfR.depth.AB.mac.snp.thin@fix)` loci

### Remove individuals with > 20% missing data and append location metadata (site, region)

```{r soft_filtering_2, echo=TRUE, message=FALSE, warning=FALSE}
# Remove INDIVIDUALS with > 20% missing data
vcfR.depth.AB.mac.snp.thin.ind <- missing_by_sample(vcfR.depth.AB.mac.snp, cutoff = 0.5)

# Upload complementary data on labels and additional aspects
my_genlight <- vcfR2genlight(vcfR.depth.AB.mac.snp.thin.ind) # turn vcf into genlight

#extra.data <- "D:/Ubuntu/Eelgrass/REFMAP/ids/Rmap_LH.txt"
#metadata <- read.delim(extra.data, stringsAsFactors = FALSE) #upload IDs

specific_data <- popmap[(popmap$id %in% indNames(my_genlight)),] # subsample IDs for those in the dataset
pop(my_genlight) <- specific_data$pop # add population to genlight
```

**After Soft Filtering**

Sample size: `r length(indNames(my_genlight))` individuals  

Variants: `r length(locNames(my_genlight))` loci


Getting an error with dartR functions. The error occurs because the R package terra (a dependency for dartR) requires the GDAL system library, which is not properly installed or configured. Here's how to resolve this:

Step 1: Install GDAL and Dependencies
On Ubuntu/Debian:
```{r engine='bash'}
sudo apt-get update
sudo apt-get install -y libgdal-dev libproj-dev gdal-bin proj-bin
```

## MAF & HWE

```{r dartr, message=FALSE, warning=FALSE, include=TRUE, echo=TRUE, results='hide'}

# dartr check1
my_genlight <- gl.compliance.check(my_genlight)

```

dartR will not install successfully. Trying new package with similar SNP filtering functionality, "melfuR"

```{r maf, echo=TRUE, message=FALSE, warning=FALSE}

# Remove LOCI with MAF < 0.05 in each of all the populations
genlight_maf <- gl.filter.maf(my_genlight, threshold = 0.05, by.pop = T, pop.limit = 4)
```

```{r hwe, echo=TRUE, message=FALSE, warning=FALSE}

# Remove LOCI with significant HWE in each of all the populations
genlight_final <- gl.filter.hwe(genlight_maf, n.pop.threshold = 4)

```

**Filtered Dataset**

Sample size: `r length(indNames(genlight_final))` individuals  

Variants: `r length(locNames(genlight_final))` loci

### 5. Saving filtered file

```{r save, echo=TRUE, message=FALSE, warning=FALSE, include=TRUE, echo=TRUE, results='hide'}

gl2genepop(genlight_final, outpath=getwd(), outfile = "./genepop_final.gen") # save the final genepop file  

## References

DeRaad, D.A. (2022), SNPfiltR: an R package for interactive and reproducible SNP filtering. Molecular Ecology Resources, 22, 2443-2453. http://doi.org/10.1111/1755-0998.13618

Knaus, Brian J., and Niklaus J. Grunwald. 2017. VCFR: a package to manipulate and visualize variant call format data in R. Molecular Ecology Resources, 17.1:44-53. http://doi.org/10.1111/1755-0998.12549

O'Leary, S. J., Puritz, J. B., Willis, S. C., Hollenbeck, C. M., & Portnoy, D. S. (2018). These aren't the loci you'e looking for: Principles of effective SNP filtering for molecular ecologists. Molecular Ecology, 27(16), 3193--3206. https://doi.org/10.1111/mec.14792
